New lower bounds for Schur and weak Schur numbers

### 2021-05-02 article/weakschur.tex
modifs

\begin{theorem}
Let \((p,q), (n,k) \in (\mathbb{N}^*)^2\). If there exists a partition of \([\![1,q]\!]\) into \(n\) weakly sum-free
subsets and a partition of \([\![1,p]\!]\) into \(k\) sum-free
subsets then there exists a partition of \([\![1,p(q+\left \lceil \frac{q}{2} \right \rceil + 1)+q]\!]\) into \(n+k\)
weakly sum-free subsets.
\end{theorem}

@@

Given that this theorem will appear as a particular case of a more general theorem after the introduction of
templates for weak Schur numbers, we only give here an intuitive explanation of the above theorem; a formal
proof can be found in the appendix.

@@

\underline{\textbf{\(\mathcal{C}\) block}} \\
In this block,  we use colors \(1, ..., n\). We colour an integer \(x\) according to its column number (written \(\tilde{\pi}(x)\), it is linked to the
projection on the first line, written \(\pi\), by the relation \(\tilde{\pi}(x) = \pi(x) - a\)). A simple solution is to colour \(x\) with the same colour
as \(\tilde{\pi}(x)\) in the weakly sum-free colouring of \([\![1,q]\!]\). As long as \(2b \leqslant a + r\) (not going two far to the right) and there
is no \(x \in \tilde{\pi}(\mathcal{C})\) such that \(2x \in \tilde{\pi}(\mathcal{C})\) (so that we do not have a sum in \(\mathcal{C}\) when
taking two numbers in the same column), the colours \(1, ..., n\) are sum-free.

In particular, taking \(w = l = \lceil \displaystyle \frac{q}{2} \rceil\) and \(r = \lfloor \displaystyle \frac{q}{2} \rfloor\) works, thus obtaining the above theorem.\\
\par
As in the previous section, we now introduce WSF-templates and the sequence \(\WS^+\) in order to generalize the above construction.

@@

Let \((a,b) \in (\mathbb{N}^*)^2\) such that \(a>b\). Define
\[ \pi_{a,b}:x \longmapsto (Id+a\mathds{1}_{ [\![0,b]\!]})(x \mod a))\]
\end{definition}

@@

Then, for every \(n \in \mathbb{N}^*\), by using on the last row the colouring \(i \longmapsto g(i - p S(n)\), we have\\
\[ \WS(n+k) \geqslant \WS^+(k+1) S(n) + b + c\]
\end{proposition}

@@

\[
\WS(n+1) \geqslant 4S(n) + 1
\]
\[
\WS(n+2) \geqslant 13S(n) + 8
\]
\[
\WS(n+3) \geqslant 42S(n) + 24
\]
\[
\WS(n+4) \geqslant 132S(n) + 26
\]

The first two inequalities were found by Rowley, they are detailed in [2]. The third inequality is optimal and was found with a SAT solver.
It uses the first sophistication explained in the previous subsection in order to add the last number in the first color.
As for the fourth inequality, it was obtained by combining an optimal SF-template of length 33 with a WSF-template of length 4.
The best template we could get with a computer search gives the inequality \(\WS(n+4) \geqslant 127 S(n) + 68\).
It was also found with the SAT solver. In order to reduce the search space, we only looked for WSF-templates of
5 colors which start with a good \(\WS(4)\) partition. However, this approach most likely prevents us from finding the best WSF-templates
as we explain in the next subsection for weakly sum-free partitions. We highly suspect that there exists more efficient WSF-templates
with \(n \geqslant 5\) colors. One may try to go over a different search space using a Monte-Carlo method, as in \cite{Bouzy2015AnAP}.
This could be the suject of a future work. Further details about the encoding as a SAT problem can be found in \cite{Heule2017}.

>>>>

\begin{theorem}
\label{theorem}
Let \((p,q), (n,k) \in (\mathbb{N}^*)^2\). If there exists a partition of \([\![1,q]\!]\) into \(n\) weakly sum-free
subsets and a partition of \([\![1,p]\!]\) into \(k\) sum-free
subsets then there exists a partition of \([\![1,p(q+\left \lceil \frac{q}{2} \right \rceil + 1)+q]\!]\) into \(n+k\)
weakly sum-free subsets.
\end{theorem}

@@

Given that this theorem will appear as a particular case of a more general theorem after the introduction of
templates for weak Schur numbers, we only give here an intuitive explanation of the above theorem; a formal
\hyperref[proof_theorem]{proof} can be found in the appendix.

@@

\underline{\textbf{\(\mathcal{C}\) block}} \\
In this block,  we use colors \(1, ..., n\). We colour an integer \(x\) according to its column number, written \(\tilde{\pi}(x)\). It is linked to the
projection on the first line, written \(\pi\), by the relation \(\tilde{\pi}(x) = \pi(x) - a\). A simple solution is to colour \(x\) with the same colour
as \(\tilde{\pi}(x)\) in the weakly sum-free colouring of \([\![1,q]\!]\). As long as \(2b \leqslant a + r\) (not going two far to the right) and there
is no \(x \in \tilde{\pi}(\mathcal{C})\) such that \(2x \in \tilde{\pi}(\mathcal{C})\) (so that we do not have a sum in \(\mathcal{C}\) when
taking two numbers in the same column), the colours \(1, ..., n\) are sum-free.

In particular, taking \(w = l = \left\lceil \displaystyle \frac{q}{2} \right\rceil\) and \(r = \left\lfloor \displaystyle \frac{q}{2} \right\rfloor\) works, thus obtaining the above theorem.\\
\par
As in the previous section, we now introduce WSF-templates and the sequence \(\WS^+\) in order to generalize the above construction.

@@

Let \((a,b) \in (\mathbb{N}^*)^2\) such that \(a>b\). Define
\[ \pi_{a,b}:x \longmapsto (Id+a\mathds{1}_{ [\![0,b]\!]})(x \mod a)\]
\end{definition}

@@

Then, for every \(n \in \mathbb{N}^*\), by using on the last row the colouring \(i \longmapsto g(i - p S(n))\), we have\\
\[ \WS(n+k) \geqslant \WS^+(k+1) S(n) + b + c\]
\end{proposition}

@@

\[
\begin{array}{r r r r r}
	WS(n + 1) & \geqslant & 4 S(n) & + & 2 \\
	WS(n + 2) & \geqslant & 13 S(n) & + & 8 \\
	WS(n + 3) & \geqslant & 42 S(n) & + & 24 \\
	WS(n + 4) & \geqslant & 132 S(n) & + & 26 \\
\end{array}
\]

The first two inequalities were found by Rowley, they are detailed in \cite{RowleyWS}. The third inequality is optimal and was found with a SAT solver.
It uses the first sophistication explained in the previous subsection in order to add the last number in the first color.
As for the fourth inequality, it was obtained by combining an optimal SF-template of length 33 with a WSF-template of length 4.
The best template we could get with a computer search gives the inequality \(\WS(n+4) \geqslant 127 S(n) + 68\).
It was also found with the SAT solver. In order to reduce the search space, we only looked for WSF-templates of
5 colors which start with a good \(\WS(4)\) partition. However, this approach most likely prevents us from finding the best WSF-templates
as we explain in the next subsection for weakly sum-free partitions. We highly suspect that there exists more efficient WSF-templates
with \(n \geqslant 5\) colors. One may try to go over a different search space using a Monte-Carlo method, as in \cite{Bouzy2015AnAP}.
This could be the suject of a future work. Further details about the encoding as a SAT problem can be found in \cite{Heule2017}.
